---
title: "final project data mining"
author: "Group 4"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---

#Required library
```{r}
# NOTE (cleanup 2026-09-27): the original file repeated library() calls in almost
# every chunk (corrplot was loaded twice). All imports are consolidated here;
# analysis code is unchanged. This Rmd was cleaned but NOT executed in this
# environment (R unavailable); run it in RStudio to verify outputs against
# reports/final-project-data-mining.docx.
library(corrplot)
library(stats)
library(factoextra)
library(ggplot2)
library(gridExtra)
library(ggplotify)
library(cluster)
library(flexclust)
library(fpc)
library(ClusterR)
library(ggpubr)
library(dplyr)
library(knitr)
library(tidyr)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
```



#Data checking and cleaning

#load data set
```{r}
# Define the column names
column_names <- c("symboling", "normalized_losses", "make", "fuel_type", "aspiration",
                  "num_of_doors", "body_style", "drive_wheels", "engine_location",
                  "wheel_base", "length", "width", "height", "curb_weight",
                  "engine_type", "num_of_cylinders", "engine_size", "fuel_system",
                  "bore", "stroke", "compression_ratio", "horsepower", "peak_rpm",
                  "city_mpg", "highway_mpg", "price")

# Read the dataset and assign the column names
data <- read.csv("imports-85.data", header = FALSE, col.names = column_names)

# Display the first few rows
head(data)

```

# Data shape 
```{r}

# ----- Data Shape -----
# Display the dimensions (rows, columns)
cat("Original Data Shape (rows, columns):", dim(data), "\n")


```


## set as numeric the numerical data  and categoric the categorical data
```{r}
# --- Convert Numerical Columns ---
data$normalized_losses  <- as.numeric(as.character(data$normalized_losses))
data$symboling          <- as.numeric(as.character(data$symboling))
data$wheel_base         <- as.numeric(as.character(data$wheel_base))
data$length             <- as.numeric(as.character(data$length))
data$width              <- as.numeric(as.character(data$width))
data$height             <- as.numeric(as.character(data$height))
data$curb_weight        <- as.numeric(as.character(data$curb_weight))
data$engine_size        <- as.numeric(as.character(data$engine_size))
data$bore               <- as.numeric(as.character(data$bore))
data$stroke             <- as.numeric(as.character(data$stroke))
data$compression_ratio  <- as.numeric(as.character(data$compression_ratio))
data$horsepower         <- as.numeric(as.character(data$horsepower))
data$peak_rpm           <- as.numeric(as.character(data$peak_rpm))
data$city_mpg           <- as.numeric(as.character(data$city_mpg))
data$highway_mpg        <- as.numeric(as.character(data$highway_mpg))
data$price              <- as.numeric(as.character(data$price))

# --- Convert Categorical Columns ---
data$make               <- as.factor(as.character(data$make))
data$fuel_type          <- as.factor(as.character(data$fuel_type))
data$aspiration         <- as.factor(as.character(data$aspiration))
data$num_of_doors       <- as.factor(as.character(data$num_of_doors))
data$body_style         <- as.factor(as.character(data$body_style))
data$drive_wheels       <- as.factor(as.character(data$drive_wheels))
data$engine_location    <- as.factor(as.character(data$engine_location))
data$engine_type        <- as.factor(as.character(data$engine_type))
data$num_of_cylinders   <- as.factor(as.character(data$num_of_cylinders))
data$fuel_system        <- as.factor(as.character(data$fuel_system))

```

# Check for Missing values 
```{r}
# Replace all occurrences of "?" with NA
data[data == "?"] <- NA

# ----- Check Missing Values -----
# Calculate the number of missing values per column
missing_counts <- sapply(data, function(x) sum(is.na(x)))
cat("Missing values in each column:\n")
print(missing_counts)


```

There are   4 missing values 

#fill missing values
```{r}
data_cleaned<-data
# Loop over all columns in the data frame
for (col_name in names(data_cleaned)) {
  # Check if the current column is numeric
  if (is.numeric(data[[col_name]])) {
    # Calculate the mean of the column while ignoring NA values
    mean_val <- mean(data_cleaned[[col_name]], na.rm = TRUE)
    
    # Replace NA values with the calculated mean
    data_cleaned[[col_name]][is.na(data_cleaned[[col_name]])] <- mean_val
  }
}
# for num_of_door since it as 2 missing values 
# ----- Fill Missing Values for the Categorical Column "num_of_doors" -----
# Check if the column exists in the data
if ("num_of_doors" %in% names(data_cleaned)) {
  # If the column is a factor, ensure four" is one of its levels
  if (is.factor(data_cleaned$num_of_doors)) {
    if (!("four" %in% levels(data_cleaned$num_of_doors))) {
      levels(data_cleaned$num_of_doors) <- c(levels(data_cleaned$num_of_doors), "four")
    }
  } else {
    # If not a factor, convert to character for imputation
    data_cleaned$num_of_doors <- as.character(data_cleaned$num_of_doors)
  }
  
  # Replace NA values with "four"
  data_cleaned$num_of_doors[is.na(data_cleaned$num_of_doors)] <- "four"
  
  # Convert the column back to factor (if desired)
  data_cleaned$num_of_doors <- as.factor(data_cleaned$num_of_doors)
}

# ----- Check Missing Values  for new data -----
# Calculate the number of missing values per column
missing_counts <- sapply(data_cleaned, function(x) sum(is.na(x)))
cat("Missing values of new data_cleaned:\n")
print(missing_counts)

```

#Check duplicated value
```{r}
# ----- Check Duplicated Values -----
# Count the number of duplicated rows in the cleaned data
num_duplicates <- sum(duplicated(data_cleaned))
cat("Number of duplicated rows:", num_duplicates, "\n")


```



# data summary
```{r}


# ----- Summary of the Data -----
cat("Summary of data:\n")
print(summary(data_cleaned))
```


# Data info 
```{r}
# ----- Data Information (Structure) -----
cat("Structure data:\n")
str(data_cleaned)


```

 


#Eda: Data visualisation



#univariate Analysis

#create numerical data
```{r}
# Create a data frame with only numeric columns from 'data'
num_data <- data_cleaned[, sapply(data_cleaned, is.numeric)]

```

#distribution for numerical data
```{r}

# Loop over each numeric column in num_data
for (col in names(num_data)) {
  p <- ggplot(num_data, aes_string(x = col)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill = "blue", color = "black", alpha = 0.6) +
    geom_density(color = "red", size = 1) +
    labs(title = paste("Distribution of", col),
         x = col,
         y = "Density") +
    theme_minimal()
  
  print(p)
}

```


# count for categorical data
```{r}

# Assuming you have a data frame 'categorical_data' that contains only categorical columns.
# If not, you can create it as follows:
 categorical_data <- data_cleaned[, !sapply(data_cleaned, is.numeric)]
# categorical_data <- data.frame(lapply(categorical_data, as.factor))

# Loop over each categorical column and create a count plot with vertical x-axis labels
for (col in names(categorical_data)) {
  p <- ggplot(categorical_data, aes_string(x = col, fill = col)) +
    geom_bar(color = "black", alpha = 0.7) +
    labs(title = paste("Count Plot for", col),
         x = col,
         y = "Count") +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
    )
  
  print(p)
}




```


#Bivariate analysis 

# box plot for Price vs Body Style
```{r}

# Create a box plot for Price vs Body Style
p <- ggplot(data_cleaned, aes(x = body_style, y = price, fill = body_style)) +
  geom_boxplot() +
  labs(title = "Price vs Body Style",
       x = "Body Style",
       y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none")

print(p)


```


#box plot for Price vs Make
```{r}

# Create a box plot for Price vs Make
p <- ggplot(data_cleaned, aes(x = make, y = price, fill = make)) +
  geom_boxplot() +
  labs(title = "Price vs Make",
       x = "Make",
       y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none")

print(p)

```


#Scatter plots for some numerical variables vs. Price
```{r}

# --- Scatter plots for numerical variables vs. Price ---

p1 <- ggplot(data_cleaned, aes(x = engine_size, y = price)) +
  geom_point(color = "blue", alpha = 0.7) +
  labs(title = "Engine Size vs Price", x = "Engine Size", y = "Price") +
  theme_minimal()

p2 <- ggplot(data_cleaned, aes(x = horsepower, y = price)) +
  geom_point(color = "darkgreen", alpha = 0.7) +
  labs(title = "Horsepower vs Price", x = "Horsepower", y = "Price") +
  theme_minimal()

p3 <- ggplot(data_cleaned, aes(x = city_mpg, y = price)) +
  geom_point(color = "purple", alpha = 0.7) +
  labs(title = "City MPG vs Price", x = "City MPG", y = "Price") +
  theme_minimal()

# --- Scatter (jitter) plots for categorical variables vs. Price ---

# Replace "Make vs Price" with "Num of Doors vs Price"
p4 <- ggplot(data_cleaned, aes(x = num_of_doors, y = price, color = num_of_doors)) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  labs(title = "Num of Doors vs Price", x = "Num of Doors", y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none")

p5 <- ggplot(data_cleaned, aes(x = body_style, y = price, color = body_style)) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  labs(title = "Body Style vs Price", x = "Body Style", y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none")

p6 <- ggplot(data_cleaned, aes(x = fuel_type, y = price, color = fuel_type)) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  labs(title = "Fuel Type vs Price", x = "Fuel Type", y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none")

# --- Display each plot individually ---
plots <- list(p1, p2, p3, p4, p5, p6)
for (p in plots) {
  print(p)
}

```


# average price for each car model (top 20)
```{r}
# Load necessary libraries


avg_prices_by_car <- data_cleaned %>%
  group_by(make) %>%
  summarise(avg_price = mean(price, na.rm = TRUE)) %>%
  arrange(desc(avg_price))

# Plot top N car models by average price
n <- 20  # Number of top car models to plot
top_car_models <- head(avg_prices_by_car, n)

# Create a horizontal bar plot
ggplot(top_car_models, aes(x = avg_price, y = reorder(make, avg_price))) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = paste("Top", n, "Car Models by Average Price"),
       x = "Average Price",
       y = "Car Model") +
  theme_minimal()

```

#corroletion for numerical data 
```{r}
# Compute the correlation matrix using the numeric data frame
cor_matrix <- cor(num_data, use = "complete.obs")

# Plot the correlation matrix as a heatmap
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45)
```


# data preprocessing /encoding
```{r}
# Identify categorical columns (non-numeric columns)
Mle_data<- data_cleaned
cat_cols <- names(Mle_data)[!sapply(Mle_data, is.numeric)]
print("Categorical columns to encode:")
print(cat_cols)

# Loop over each categorical column and encode it as numeric
for (col in cat_cols) {
  Mle_data[[col]] <- as.numeric(as.factor(Mle_data[[col]]))
}

```


#Unsupervised learning 

# Outlier Detection 
```{r}
# Outlier Detection using IQR Rule
auto_norm <- as.data.frame(scale(Mle_data))

vars <- names(auto_norm)
for (var in vars) {
  Q1 <- quantile(auto_norm[[var]], 0.25, na.rm = TRUE)
  Q3 <- quantile(auto_norm[[var]], 0.75, na.rm = TRUE)
  IQR_val <- IQR(auto_norm[[var]], na.rm = TRUE)
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  outlier_count <- length(which(auto_norm[[var]] < lower_bound | auto_norm[[var]] > upper_bound))
  cat(paste(var, "outliers count:", outlier_count, "\n"))
}
```

#Clusterability Assessment
```{r}

# Use the Hopkins statistic from factoextra to assess clustering tendency
hopkins_stat <- get_clust_tendency(auto_norm, n = 20, graph = FALSE)$hopkins_stat
cat("Hopkins statistic:", hopkins_stat, "\n")
# (A value near 0.5 indicates clusterability, while values near 1 suggest randomness.)

```

#Average number of cluster
```{r}

### Average of number of cluster
k <- fviz_nbclust(auto_norm, FUNcluster = kmeans, method = "silhouette") + ggtitle("K-Means")
e <- fviz_nbclust(auto_norm, FUNcluster = clara, method = "silhouette") + ggtitle("CLARA")
grid.arrange(k, e,  ncol = 2)
```

# kmeans 
```{r}
#Determine the Optimal Number of Clusters
# Using the elbow method, let’s confirm the optimal number of clusters to be used for K-Means Clustering Analysis.
Optimal_Clusters_KMeans(auto_norm, max_clusters=10, plot_clusters = TRUE)



# K-Means Clustering

set.seed(123)
kmeans_result <- kmeans(auto_norm, centers = 2, nstart = 25)
# Visualize K-Means clusters
clus_kmeans <- fviz_cluster(list(data = auto_norm, cluster = kmeans_result$cluster),
                            ellipse.type = "norm", geom = "point", stand = FALSE,
                            palette = "jco", ggtheme = theme_classic())
# Silhouette plot for K-Means
df_sil_kmeans <- silhouette(kmeans_result$cluster, dist(auto_norm))
sil_kmeans <- fviz_silhouette(df_sil_kmeans)
grid.arrange(clus_kmeans, sil_kmeans, ncol = 2)


stat_km2 <- cclust(auto_norm, 2, dist="euclidean")
stripes(stat_km2)

```




#clara
```{r}

# CLARA Clustering

set.seed(123)
clara_result <- clara(auto_norm, k = 3, samples = 5)

# Visualize CLARA clusters using factoextra
clus_clara <- fviz_cluster(list(data = auto_norm, cluster = clara_result$clustering),
                           ellipse.type = "norm", geom = "point", stand = FALSE,
                           palette = "jco", ggtheme = theme_classic())

# Create the silhouette plot for CLARA clustering
df_sil_clara <- silhouette(clara_result$clustering, dist(auto_norm))
sil_clara <- fviz_silhouette(df_sil_clara)

# Arrange the cluster plot and silhouette plot side by side
grid.arrange(clus_clara, sil_clara, ncol = 2)


```


#PCA 
```{r}

#  Perform PCA Analysis

pca_res <- prcomp(auto_norm, center = TRUE, scale. = FALSE)

# Get a summary of the PCA results
pca_summary <- summary(pca_res)
print(pca_summary)

# Create a table of the PCA component loadings
pca_loadings <- pca_res$rotation
kable(round(pca_loadings, 3), caption = "PCA Component Loadings")


#  Create a Scree Plot

#  Calculate eigenvalues
eigenvalues <- pca_res$sdev^2

#  Create a data frame, making PC a factor with explicit levels
scree_data <- data.frame(
  PC = factor(paste0("PC", 1:length(eigenvalues)), 
              levels = paste0("PC", 1:length(eigenvalues))),
  Eigenvalue = eigenvalues
)

#  Plot the scree plot, ensuring the x-axis follows the factor order
scree_plot <- ggplot(scree_data, aes(x = PC, y = Eigenvalue)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_line(aes(group = 1), color = "red", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(title = "Scree Plot", x = "Principal Components", y = "Eigenvalue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))



#  Create a PCA Biplot

biplot_plot <- fviz_pca_biplot(pca_res,
                               repel = TRUE,
                               col.var = "blue",  # Variables color
                               col.ind = "gray",  # Individuals color
                               title = "PCA Biplot")

# Display the Plots Side by Side

grid.arrange(scree_plot, biplot_plot, ncol = 2)

```


#supervised learning 

## Linear regression 
```{r}
#install.packages("caret")
```

```{r}

set.seed(123)  # for reproducibility

#  Split the Data into Training (70%) and Testing (30%) Sets ---
train_index <- createDataPartition(auto_norm$price, p = 0.7, list = FALSE)
train_data <- Mle_data[train_index, ]
test_data <- Mle_data[-train_index, ]
```



```{r}
#  Fit a Linear Regression Model on the Training Data
model_1 <- lm(price ~ ., data = train_data)

# Print the full model summary (R², coefficients, p-values, F-statistics, etc.)
summary_model <- summary(model_1)
print(summary_model)
```



```{r}
#  Predict on the Test Set and Calculate MAE
predictions <- predict(model_1, newdata = test_data)
mae_val <- mean(abs(predictions - test_data$price))
cat("Linear Regression MAE:", round(mae_val, 2), "\n")

#  Perform 10-Fold Cross Validation for the Model 
control <- trainControl(method = "cv", number = 10)
set.seed(123)
model_cv <- train(price ~ ., data = train_data, method = "lm", trControl = control)
cv_score <- model_cv$results$Rsquared[1]
cat("Cross Validation R-squared score:", round(cv_score, 2), "\n")
```




```{r}
#  Plot Actual vs Predicted Price on the Test Set 
plot_data <- data.frame(Actual = test_data$price, Predicted = predictions)
p <- ggplot(plot_data, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted Price",
       x = "Actual Price",
       y = "Predicted Price") +
  theme_minimal()

print(p)

```




```{r}
# Create a data frame with actual values, predicted values, and differences for the first 10 test observations
result_table <- data.frame(
  Car = 1:10,
  Actual = test_data$price[1:10],
  Predicted = predictions[1:10]
)

# Compute the absolute differences for each car
result_table$Difference <- abs(result_table$Actual - result_table$Predicted)

# Print the table
print(result_table)



```


## Decision Tree 
```{r}
# Load necessary libraries

# Fit the Decision Tree Model 
# Fit a regression tree using rpart with method "anova" for regression
model_tree <- rpart(price ~ ., data = train_data, method = "anova")

# Print model details and complexity parameter table
print(model_tree)
printcp(model_tree)
```

```{r}
pred_tree <- predict(model_tree, newdata = test_data)
plot_data <- data.frame(Actual = test_data$price, Predicted = pred_tree)
p_tree <- ggplot(plot_data, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Decision Tree: Actual vs Predicted Price",
       x = "Actual Price",
       y = "Predicted Price") +
  theme_minimal()

print(p_tree)
```


```{r}

# Calculate Mean Absolute Error (MAE) for the tree model
mae_tree <- mean(abs(test_data$price - pred_tree))
cat("Decision Tree MAE: ", round(mae_tree, 2), "\n")
```



```{r}
# Compute the sum of squared errors (SSE) and total sum of squares (SST)
SSE <- sum((test_data$price - pred_tree)^2)
SST <- sum((test_data$price - mean(test_data$price))^2)
r_squared <- 1 - SSE/SST

cat("Decision Tree R-squared:", round(r_squared, 2), "\n")

```

#random forest 
```{r}
#install.packages("randomForest")
```

```{r}
# Load necessary libraries


set.seed(123)  # For reproducibility

#  Fit the Random Forest Model 
# Using 500 trees; adjust ntree as needed
model_rf <- randomForest(price ~ ., data = train_data, ntree = 500, importance = TRUE)

#  Predict on the Test Set 
pred_rf <- predict(model_rf, newdata = test_data)

# Calculate MAE and R-squared 
# Mean Absolute Error (MAE)
mae_rf <- mean(abs(test_data$price - pred_rf))

# R-squared Calculation
SSE <- sum((test_data$price - pred_rf)^2)
SST <- sum((test_data$price - mean(test_data$price))^2)
r_squared_rf <- 1 - SSE/SST

cat("Random Forest MAE:", round(mae_rf, 2), "\n")
cat("Random Forest R-squared:", round(r_squared_rf, 2), "\n")
```


```{r}
# Plot Actual vs Predicted Prices 
plot_data_rf <- data.frame(Actual = test_data$price, Predicted = pred_rf)

p_rf <- ggplot(plot_data_rf, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Random Forest: Actual vs Predicted Price",
       x = "Actual Price",
       y = "Predicted Price") +
  theme_minimal()

print(p_rf)

```

